---
output:
  github_document:
    md_extensions: -smart
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# vigiar

<!-- badges: start -->
[![R-CMD-check](https://github.com/santosry/vigiar/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/santosry/vigiar/actions/workflows/R-CMD-check.yaml)
[![lint](https://github.com/santosry/vigiar/actions/workflows/lint.yaml/badge.svg)](https://github.com/santosry/vigiar/actions/workflows/lint.yaml)
[![pkgdown](https://github.com/santosry/vigiar/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/santosry/vigiar/actions/workflows/pkgdown.yaml)
[![test-coverage](https://github.com/santosry/vigiar/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/santosry/vigiar/actions/workflows/test-coverage.yaml)
[![online-canary](https://github.com/santosry/vigiar/actions/workflows/online-canary.yaml/badge.svg)](https://github.com/santosry/vigiar/actions/workflows/online-canary.yaml)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![R >= 4.1.0](https://img.shields.io/badge/R-%3E%3D%204.1.0-blue.svg)](https://www.r-project.org/)
<!-- badges: end -->

`vigiar` is an R package for downloading, processing, validating, auditing,
and documenting public data from Brazil's VIGIAR environmental health
surveillance dashboard. The first robust workflow targets PM2.5 and the 92
municipalities of Rio de Janeiro state.

The package is deliberately conservative: it does not merely filter rows for
RJ. It measures whether the RJ download is complete, identifies missing
municipalities, warns about possible API truncation, and keeps reproducibility
metadata on the returned data.

The generic `vigiar_baixar()` function does not restrict geography by default.
RJ scope is explicit through `vigiar_baixar_rj()`.

## Installation

```{r}
# install.packages("remotes")
remotes::install_github("santosry/vigiar")
```

Core dependencies are `httr2`, `jsonlite`, `tibble`, `dplyr`, `cli`, and
`openssl`. Vignettes use `knitr` and `rmarkdown`. The exploratory RJ plot uses
`ggplot2` only when it is installed.

## Quick Start

```{r}
library(vigiar)

vigiar_conectar()

pm25_rj <- vigiar_baixar_rj("df_anual", validar_cobertura = TRUE)
cobertura <- vigiar_rj_cobertura(pm25_rj)
print(cobertura)

pm25_rj <- process_pm25(pm25_rj, tipo = "anual")
diag <- vigiar_diagnosticar_serie(pm25_rj, escopo = "rj")
vigiar_relatorio_diagnostico(diag)

snap <- vigiar_snapshot(dados = pm25_rj, tabela = "df_anual")
vigiar_salvar_snapshot(snap, "pm25_rj_snapshot.rds")

vigiar_desconectar()
```

## Evidence Vocabulary

The package uses these terms deliberately:

- **downloaded**: the endpoint returned a response that the parser accepted;
- **filtered**: rows were restricted to a requested scope;
- **covered**: expected spatial units appear at least once;
- **complete**: every unit in an explicit expected spatial-temporal grid appears;
- **verified**: schema, expected domain, and response evidence were checked;
- **audited**: those checks and their outcomes were recorded in a structured report;
- **reproducible**: data, schema, metadata, software version, and checksums were archived.

These terms are not synonyms. A successful download can be partial, a 92/92
spatial coverage result can have missing municipality-month cells, and a
reproducible snapshot can preserve incomplete source data.

## Downloading Rio de Janeiro Municipality Data

Use `vigiar_baixar_rj()` when the analysis scope is Rio de Janeiro. It downloads
the requested VIGIAR table, filters against the internal registry of the 92 RJ
municipalities, normalizes 6- and 7-digit IBGE municipality codes to the package
standard, and attaches coverage metadata.

```{r}
library(vigiar)

vigiar_conectar()

pm25_rj <- vigiar_baixar_rj("df_anual", validar_cobertura = TRUE)

cobertura <- vigiar_rj_cobertura(pm25_rj)
print(cobertura)

ausentes <- vigiar_rj_municipios_ausentes(pm25_rj)
print(ausentes)

pm25_rj <- process_pm25(pm25_rj, tipo = "anual")

diag <- vigiar_diagnosticar_serie(pm25_rj, escopo = "rj")
vigiar_relatorio_diagnostico(diag)

vigiar_desconectar()
```

When the research question requires all 92 municipalities, set
`exigir_completo = TRUE` or `require_complete = TRUE`. The function stops with a
clear error if any expected RJ municipality is absent or if possible API
truncation prevents the package from guaranteeing completeness.

```{r}
pm25_rj <- vigiar_baixar_rj(
  "df_anual",
  validar_cobertura = TRUE,
  require_complete = TRUE
)
```

Some VIGIAR tables may naturally contain fewer than 92 municipalities for a
given table, year, or month. A missing municipality can indicate limited source
availability, a schema change, an inappropriate filter, API truncation, or a
true absence from the source table. The package reports this honestly, but the
researcher must decide whether the resulting data are epidemiologically fit for
the intended analysis.

Useful RJ metadata are stored as attributes:

```{r}
attr(pm25_rj, "vigiar_uf")
attr(pm25_rj, "vigiar_rj_n_municipios")
attr(pm25_rj, "vigiar_rj_n_esperado")
attr(pm25_rj, "vigiar_rj_cobertura_pct")
attr(pm25_rj, "vigiar_rj_municipios_ausentes")
attr(pm25_rj, "vigiar_possivel_truncamento")
```

## Municipality Code Standard

`vigiar` uses the 6-digit IBGE municipality code internally. The RJ registry
also includes `codigo_ibge_7`, because some public sources use the official
7-digit code with a check digit.

```{r}
vigiar_rj_municipios()
```

The internal normalizer accepts integer, numeric, and character inputs such as
`330455`, `3304557`, `"330455"`, and `" 330455 "`. Values that cannot be safely
normalized return `NA`.

The internal RJ registry is locked against offline reference fixtures generated
from official sources:

- `inst/extdata/municipios_ibge_reference.csv` is a versioned national IBGE
  municipality reference derived from the IBGE Localidades API; exactly 92 rows
  belong to RJ.
- `inst/extdata/municipios_ibge_reference_metadata.json` records its checksum,
  retrieval date, source URL, and row counts.
- `inst/extdata/rj_regioes_saude_sesrj_reference.csv` locks the 9 SES-RJ health
  regions used by the package.
- `inst/extdata/rj_official_sources.csv` records the official source URLs and
  access date used for these fixtures.

For municipality-specific work, prefer code-based downloads instead of filtering
by municipality name. This avoids fragile joins caused by spelling, accents, or
source-specific name variants.

```{r}
campos <- vigiar_baixar_municipio("df_anual", codigo_ibge = 330100)

attr(campos, "vigiar_codigo_ibge_6")
attr(campos, "vigiar_codigo_ibge_7")
attr(campos, "vigiar_municipio")
attr(campos, "vigiar_regiao_saude")
```

## Coverage Diagnostics

`vigiar_rj_cobertura()` returns a tibble with:

- expected municipality count (`n_municipios_esperados = 92`);
- observed municipality count;
- coverage percentage;
- absent municipality names;
- incomplete SES-RJ health regions;
- a completeness flag;
- a possible truncation flag.

Coverage can be calculated overall or by time unit:

```{r}
vigiar_rj_cobertura(pm25_rj)
vigiar_rj_cobertura(pm25_rj, por = "ano")
vigiar_rj_cobertura(pm25_rj, por = "mes")
vigiar_rj_cobertura(pm25_rj, por = "ano_mes")
vigiar_rj_cobertura(pm25_rj, por = "regiao_saude")
```

`por = "macrorregiao"` and `macrorregiao_saude` remain compatibility aliases.
The canonical scientific term for the package's nine SES-RJ groups is **health
region**, represented by `regiao_saude`.

For table-aware completeness, use `vigiar_rj_completude_tabela()`. It applies the
expected grid for common VIGIAR tables: municipality x year for annual tables,
municipality x year for `pop`, municipality x year x month for monthly PM2.5
tables, and municipality x year x month for daily-reference tables when `ano`
and `mes` are available.

```{r}
vigiar_rj_completude_tabela(pm25_rj, tabela = "df_anual")

vigiar_rj_completude_tabela(
  pm25_rj,
  tabela = "df_anual",
  anos_esperados = 2018:2023,
  require_complete = TRUE
)
```

Without `anos_esperados`, `meses_esperados`, `periodo_inicio`/`periodo_fim`, or
an official domain, the function may infer the interval between observed
endpoints. That result is `complete_within_inferred_domain`, not proof that the
researcher's expected study period is complete.

`vigiar_diagnosticar_serie(..., escopo = "rj")` uses these coverage checks and
adds clear messages such as complete RJ coverage, partial RJ coverage, low
coverage in a specific year, no valid RJ municipality code, or possible API
truncation.

## Reproducible Online RJ Audit

Use `vigiar_auditar_rj_online()` before formal scientific use or before
archiving a release. It downloads the requested RJ tables, computes checksum and
schema hash, evaluates table-grain completeness, lists absent municipalities,
records temporal gaps, verifies parser structure, flags possible Power BI
truncation, and writes CSV, JSON, and RDS artifacts under
`data-raw/rj-download-completeness-output/`.

```{r}
vigiar_conectar()

audit <- vigiar_auditar_rj_online(
  tabelas = c("df_anual", "df_mensal", "df_dias", "pop"),
  salvar = TRUE,
  require_complete = FALSE
)

audit[, c(
  "tabela",
  "n_rows",
  "checksum",
  "completeness_grade",
  "n_incomplete_groups",
  "parser_status",
  "possivel_truncamento",
  "conclusion"
)]

vigiar_desconectar()
```

When `require_complete = TRUE`, the audit stops if any table is partial,
truncated, has parser issues, lacks a municipality code, lacks a schema hash,
or otherwise fails. It also requires an explicit expected temporal domain. This
is intentionally stricter than checking that `nrow(dados) > 0`.

```{r}
years <- 2018:2023

strict_audit <- vigiar_auditar_rj_online(
  tabelas = c("df_anual", "df_mensal", "df_dias", "pop"),
  dominios_esperados = list(
    df_anual = list(anos_esperados = years),
    df_mensal = list(anos_esperados = years, meses_esperados = 1:12),
    df_dias = list(anos_esperados = years, meses_esperados = 1:12),
    pop = list(anos_esperados = years)
  ),
  salvar = TRUE,
  require_complete = TRUE
)
```

The objective conclusion can be `complete`, `complete_within_inferred_domain`,
`complete_within_observed_domain`, `partial`, `truncated`, `schema_changed`,
`schema_unverified`, or `failed`. Plain `complete` is reserved for an explicit
expected domain, complete RJ panel, matching critical schema, and no truncation
evidence or parser issues. It does not prove causal validity, exposure
measurement validity, model adequacy, or epidemiological identifiability.

Truncation evidence is recorded separately as `no_evidence`, `possible`,
`probable`, `confirmed`, or `unknown`. The raw response is assessed before any
client-side UF filter, so filtering cannot erase a response-limit signal.
Parser evidence is recorded as `parser_status`, `parser_issue_count`, and
`parser_issues`. Power BI may mix dictionary indices with inline text after a
dictionary reaches its cap; valid inline values are preserved, while actual
row-width, mask, or dictionary-index violations prevent a `complete` result.

## Processing

```{r}
pm25 <- vigiar_baixar_rj("df_anual") |>
  process_pm25(tipo = "anual")

population <- vigiar_baixar_rj("pop") |>
  process_populacao_exposta()

health <- vigiar_baixar_rj("tb_muni") |>
  process_indicadores_saude(agregacao = "municipio")
```

Processed tibbles keep package metadata and S3 classes such as `vigiar_pm25`,
`vigiar_health`, and `vigiar_population`.

## Snapshots, Cache, and Audit

```{r}
vigiar_cache_dir("~/.cache/vigiar")
pm25 <- vigiar_baixar_rj("df_anual", usar_cache = TRUE)

snapshot <- vigiar_snapshot(dados = pm25, tabela = "df_anual")
vigiar_verificar_snapshot(snapshot)

audit <- vigiar_auditar(pm25, tabela = "df_anual")
vigiar_exportar_auditoria(audit, "audit-pm25-rj.json")
```

Snapshots preserve RJ metadata, including coverage and truncation attributes.
Canonicalization version 2 stores separate data, schema, and metadata checksums.
`vigiar_checksum(mode = "canonical")` ignores row/column order and equivalent
factor/character or integer/double representations while preserving full
numeric precision; `mode = "ordered"` retains table order as part of identity.

Schema locks can be used to detect dashboard changes before an analysis is
reused. `vigiar_esquema_verificar_critico()` checks the locked critical columns
used by the RJ PM2.5, municipality, coordinate, and population workflows.

```{r}
vigiar_esquema_lock("vigiar_schema_lock.json")
vigiar_esquema_verificar("vigiar_schema_lock.json", error = TRUE)
vigiar_esquema_verificar_critico(error = TRUE)
```

The manual validation script `data-raw/check-rj-download-completeness.R` writes
timestamped release reports for `df_muni`, `df_anual`, `df_mensal`, `df_dias`,
and `df_dias_conama`, including checksums, table-grain coverage tables, and
missing-municipality, missing-year, and missing-month reports under
`data-raw/rj-download-completeness-output/`.
Set `VIGIAR_VALIDATION_RELEASE` before running the script to bind the output to
a formal release. Generated reports are intended to be archived with a release
or local validation record, not committed as package data.

For strict runs, set `VIGIAR_EXPECTED_YEARS` or both
`VIGIAR_EXPECTED_PERIOD_START` and `VIGIAR_EXPECTED_PERIOD_END`. The script
refuses `VIGIAR_REQUIRE_COMPLETE=true` without an explicit temporal domain.

The optional online test is also an audit, not only a smoke test. With
`VIGIAR_RUN_ONLINE_TESTS=true`, it downloads `df_anual`, `df_mensal`, and
`df_dias`, and `pop`, writes a manifest with timestamp, row count, checksum,
schema and verification states, missing
municipalities, missing years, and missing months, and validates the expected
table grain for each table. Set `VIGIAR_ONLINE_REPORT_DIR` to choose the output
directory and set `VIGIAR_REQUIRE_ONLINE_COMPLETENESS=true` for formal release
checks that must fail when table-grain completeness or truncation is detected.
The scheduled `online-canary` workflow archives these artifacts independently
of pull-request checks. Offline coverage is a separate CI gate at 70%, with a
Cobertura XML report and RDS evidence artifact.

## API Lifecycle

`vigiar_api_status()` classifies every exported symbol as `stable`,
`experimental`, or `deprecated`. Internal dot-prefixed helpers are not exported.
The language and compatibility policy is documented in
[`LANGUAGE_POLICY.md`](LANGUAGE_POLICY.md).

## Exploratory Plot

```{r}
pm25 <- vigiar_baixar_rj("df_anual", processar = TRUE, tipo = "anual")

vigiar_plot_pm25_rj(pm25, por = "ano")
vigiar_plot_pm25_rj(pm25, por = "macrorregiao") # compatibility alias
```

This plot is exploratory. It does not download, process, or repair data
silently.

## Tables

| Table | Main content |
| --- | --- |
| `df_anual` | Annual PM2.5 by municipality |
| `df_mensal` | Monthly PM2.5 by municipality |
| `df_dias` | Days above the WHO daily PM2.5 reference |
| `df_dias_conama` | Days above the CONAMA PM2.5 reference |
| `pop` | Population exposure |
| `df_muni` | Municipality registry |
| `tb_brasil` | Brazil-level health indicators |
| `tb_uf` | UF-level health indicators |
| `tb_muni` | Municipality-level health indicators |
| `tb_fracao` | Attributable fraction estimates |
| `tb_quartis` | Indicator quartiles |
| `df_indoor` | Indoor solid-fuel exposure |
| `df_indoor_desfecho` | Indoor exposure health outcomes |
| `medidas` | Calculated dashboard measures |

Run `vigiar_info()` for the live table catalogue after connecting.

## Scientific Cautions

- VIGIAR data are public, aggregated surveillance data.
- PM2.5 may be estimated or modelled, not measured at every ground location.
- A dataset can be available and still be incomplete for RJ.
- Spatial coverage must be checked before municipality or health-region claims.
- The Power BI API may truncate large responses.
- Tables without municipality codes cannot prove 92/92 RJ coverage.
- Short time series and missing months limit trend interpretation.
- Ecological associations do not establish individual-level causality.
- This package prepares and audits data. It does not implement or validate
  causal inference, GAM, DLNM, relative-risk, or machine-learning models.

## Citation

> Santos, R. (2026). vigiar: VIGIAR Environmental Health Data for Rio de
> Janeiro. R package version 0.7.1.9000.
> https://github.com/santosry/vigiar

## License

MIT. Downloaded public data remain under the responsibility and terms of the
Brazilian Ministry of Health / VIGIAR source.
