---
title: "Monthly PM2.5 for Rio de Janeiro"
author: "Ryan Santos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Monthly PM2.5 for Rio de Janeiro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Purpose

`vigiar_baixar_pm25_mensal_rj()` is the dedicated entry point for monthly PM2.5
means from the VIGIAR `df_mensal` table. It does more than return rows. It:

- restricts data to the official registry of 92 RJ municipalities;
- normalizes municipality identifiers to six-digit IBGE codes;
- preserves official seven-digit codes for interoperability;
- processes the source value into `pm25_media` in `ug/m3`;
- creates a first-day monthly `periodo` date;
- measures municipality by year by month completeness;
- preserves parser, schema, query, checksum, partition, and truncation evidence;
- can stop instead of returning data when completeness cannot be demonstrated.

Successful retrieval, 92/92 overall coverage, and a complete monthly panel are
different claims. A table can contain every RJ municipality at least once and
still omit one municipality in one month.

## Download Modes

Use exploratory mode while discovering a table or checking source availability:

```{r, eval = FALSE}
vigiar_conectar()

monthly <- vigiar_baixar_pm25_mensal_rj(
  anos = 2024,
  meses = 1:12,
  validar_completude = TRUE
)

vigiar_desconectar()
```

Use strict mode when a scientific workflow requires a declared complete panel:

```{r, eval = FALSE}
vigiar_conectar()

monthly <- vigiar_baixar_pm25_mensal_rj(
  anos_esperados = 2020:2024,
  meses_esperados = 1:12,
  require_complete = TRUE
)

vigiar_desconectar()
```

Strict mode requires an explicit expected temporal domain. It downloads
server-side year-month partitions and fails if any partition fails, any expected
municipality-month cell is absent, the key is duplicated, PM2.5 values are
impossible, the parser or critical schema does not pass, or truncation evidence
is present. Observed data are never allowed to define their own strict boundary.

## Offline Complete Example

The following data are simulated. No network request is made while building
this article.

```{r}
library(vigiar)

rj <- vigiar_rj_municipios()
periods <- expand.grid(
  ano = 2024L,
  mes = 1:3,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

monthly <- merge(
  rj[c("codigo_ibge_6", "codigo_ibge_7", "municipio", "regiao_saude")],
  periods,
  by = NULL
)
monthly$cod_municipio <- monthly$codigo_ibge_6
monthly$pm25_media <- 8 +
  (as.integer(monthly$codigo_ibge_6) %% 17) / 10 +
  monthly$mes / 10
monthly$periodo <- as.Date(sprintf(
  "%04d-%02d-01", monthly$ano, monthly$mes
))

attr(monthly, "vigiar_parser_status") <- "pass"
attr(monthly, "vigiar_schema_status") <- "pass"
attr(monthly, "vigiar_truncation_status") <- "no_evidence"
```

Define the expected grid explicitly:

```{r}
completeness <- vigiar_rj_completude_tabela(
  monthly,
  tabela = "df_mensal",
  anos_esperados = 2024L,
  meses_esperados = 1:3
)

completeness[, c(
  "ano", "mes", "n_municipios_presentes",
  "n_municipios_esperados", "n_ausentes", "completo"
)]
```

Run the descriptive audit without downloading again:

```{r}
analysis <- vigiar_analisar_pm25_mensal_rj(
  monthly,
  completude = completeness,
  schema_status = "pass"
)

analysis
analysis$summary
analysis$temporal_domain
analysis$by_year_month
analysis$quality
analysis$quality_by_year_month
```

## Output Contract

The monthly tibble uses these canonical fields:

| Field | Contract |
|---|---|
| `codigo_ibge_6` | Six-digit internal municipality code |
| `codigo_ibge_7` | Official seven-digit interoperability code when available |
| `municipio` | Official municipality name from the internal RJ registry |
| `regiao_saude` | Canonical SES-RJ health region |
| `ano`, `mes` | Integer temporal identifiers |
| `periodo` | First day of the month as `Date` |
| `pm25_media` | Monthly PM2.5 mean in `ug/m3` |

Important attributes include:

| Attribute | Evidence |
|---|---|
| `vigiar_rj_completude` | Expected municipality-year-month grid |
| `vigiar_monthly_analysis` | Descriptive summaries, quality, and conclusion |
| `vigiar_partition_report` | Attempt, retry, row count, and status per partition |
| `vigiar_parser_status` | Structural Power BI response assessment |
| `vigiar_schema_status` | Critical-schema compatibility |
| `vigiar_truncation_status` | `no_evidence`, `possible`, `probable`, `confirmed`, or `unknown` |
| `vigiar_schema_hash` | Download-time schema identity when available |
| `vigiar_download_timestamp` | Retrieval timestamp |

The analysis object exposes structured fields. Code should inspect
`overall_status`, `conclusion`, and the evidence tables instead of parsing its
printed text.

## Detecting One Missing Municipality-Month

Remove Campos dos Goytacazes from February while retaining it in the other
months:

```{r}
partial <- monthly[
  !(monthly$codigo_ibge_6 == "330100" & monthly$mes == 2L),
  ,
  drop = FALSE
]
attr(partial, "vigiar_parser_status") <- "pass"
attr(partial, "vigiar_schema_status") <- "pass"
attr(partial, "vigiar_truncation_status") <- "no_evidence"

partial_completeness <- vigiar_rj_completude_tabela(
  partial,
  tabela = "df_mensal",
  anos_esperados = 2024L,
  meses_esperados = 1:3
)
partial_analysis <- vigiar_analisar_pm25_mensal_rj(
  partial,
  completude = partial_completeness,
  schema_status = "pass"
)

partial_completeness[, c(
  "ano", "mes", "n_municipios_presentes", "n_ausentes", "completo"
)]
partial_analysis$missing_municipalities[
  partial_analysis$missing_municipalities$codigo_ibge_6 == "330100",
]
partial_analysis$conclusion
```

Overall municipal coverage may still be 92/92 because Campos appears in January
and March. The table-grain conclusion is nevertheless `partial`, which is the
scientifically relevant result for a monthly panel.

## Quality Evidence

`vigiar_analisar_pm25_mensal_rj()` reports:

- missing PM2.5 values;
- negative values;
- suspicious zeros;
- values above the configurable plausibility threshold;
- invalid municipality codes;
- invalid year or month values;
- duplicated municipality-year-month keys;
- long missing blocks and abrupt changes between consecutive months;
- missing municipalities for each expected period;
- failed download partitions;
- parser, schema, and truncation states.

Values above `limite_plausibilidade` are flagged for unit and source review. The
function does not silently remove, replace, winsorize, or impute observations.

## Reproducible Release Evidence

For a multi-table release audit, use `vigiar_auditar_rj_online()` with explicit
expected domains. The audit writes CSV, JSON, and RDS evidence containing the
package version, commit SHA, timestamp, dimensions, canonical checksum, schema
hash, coverage, incomplete periods, missing municipalities, parser status,
truncation status, and objective conclusion.

```{r, eval = FALSE}
years <- 2010:2024

audit <- vigiar_auditar_rj_online(
  tabelas = c("df_anual", "df_mensal", "df_dias", "pop"),
  dominios_esperados = list(
    df_anual = list(anos_esperados = years),
    df_mensal = list(anos_esperados = years, meses_esperados = 1:12),
    df_dias = list(anos_esperados = years, meses_esperados = 1:12),
    pop = list(anos_esperados = years)
  ),
  salvar = TRUE,
  require_complete = TRUE
)
```

The manual release runner is `data-raw/check-rj-download-completeness.R`.
Generated evidence is written under
`data-raw/rj-download-completeness-output/` and should be archived with the
corresponding release rather than installed as package data.

## Scientific Interpretation

Strict completeness is evidence that the declared municipality-time grid was
retrieved with passing structural checks at one timestamp. It does not prove:

- that PM2.5 represents ground-monitor measurements in every municipality;
- absence of exposure measurement error;
- individual-level exposure or risk;
- causal validity;
- adequacy of GAM, DLNM, relative-risk, predictive, or machine-learning models;
- future availability or schema stability of the external Power BI source.

Use these outputs as retrieval and data-quality evidence. Epidemiologic fitness
still depends on the study design, exposure definition, outcome quality,
confounding control, uncertainty analysis, and ecological interpretation.
