---
title: "Verifiable Rio de Janeiro Downloads"
author: "Ryan Santos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Verifiable Rio de Janeiro Downloads}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Why Completeness Matters

For Rio de Janeiro, the package expects 92 municipalities. A table filtered to
RJ is not automatically complete. Completeness must be measured against the
internal registry and, when time columns exist, by year or by year-month.

The executable examples below use simulated data so this vignette builds
offline.

A response can be downloaded and filtered without being complete. Overall
92/92 coverage also does not prove that every municipality is present in every
expected year or month. `vigiar` reports spatial coverage, temporal-domain
status, panel completeness, schema status, and response truncation separately.

```{r}
library(vigiar)

rj <- vigiar_rj_municipios()
code_col <- if ("codigo_ibge_6" %in% names(rj)) "codigo_ibge_6" else "codigo_ibge"

coverage_fun <- function(dados, por = "geral") {
  if (exists("vigiar_rj_cobertura", mode = "function")) {
    return(vigiar_rj_cobertura(dados, por = por))
  }

  if (por == "ano" && "ano" %in% names(dados)) {
    years <- sort(unique(dados$ano))
    return(do.call(rbind, lapply(years, function(y) {
      out <- coverage_fun(dados[dados$ano == y, ], por = "geral")
      out$nivel <- "ano"
      out$ano <- y
      out
    })))
  }

  present <- unique(dados$cod_municipio)
  present <- present[present %in% rj[[code_col]]]
  data.frame(
    nivel = por,
    ano = NA_integer_,
    mes = NA_integer_,
    n_municipios_presentes = length(present),
    n_municipios_esperados = nrow(rj),
    cobertura_pct = round(100 * length(present) / nrow(rj), 1),
    n_ausentes = nrow(rj) - length(present),
    completo = length(present) == nrow(rj),
    stringsAsFactors = FALSE
  )
}

missing_fun <- function(dados) {
  if (exists("vigiar_rj_municipios_ausentes", mode = "function")) {
    return(vigiar_rj_municipios_ausentes(dados))
  }
  present <- unique(dados$cod_municipio)
  rj[!rj[[code_col]] %in% present, ]
}

head(rj)
```

## Simulated Complete Coverage

```{r}
pm25_complete <- data.frame(
  cod_municipio = rj[[code_col]],
  ano = 2022L,
  pm25_media_anual = 12,
  stringsAsFactors = FALSE
)

coverage_fun(pm25_complete)
```

## Simulated Partial Coverage

```{r}
pm25_partial <- pm25_complete[1:10, ]

cov_partial <- coverage_fun(pm25_partial)
cov_partial

missing_fun(pm25_partial)
```

## Coverage by Year

```{r}
pm25_years <- rbind(
  transform(pm25_complete, ano = 2020L),
  transform(pm25_partial, ano = 2021L)
)

coverage_fun(pm25_years, por = "ano")
```

## Table-Aware Completeness

`vigiar_rj_completude_tabela()` applies the expected municipal grid for common
VIGIAR tables. For `df_anual`, the expected grid is municipality by year. For
`pop`, the expected grid is also municipality by year. For `df_mensal`,
`df_dias`, and `df_dias_conama`, the expected grid is municipality by year by
month when `ano` and `mes` are available.

```{r}
if (exists("vigiar_rj_completude_tabela", mode = "function")) {
  vigiar_rj_completude_tabela(
    pm25_years,
    tabela = "df_anual",
    anos_esperados = 2020:2022
  )
}
```

```{r}
pm25_months <- merge(
  pm25_complete,
  data.frame(mes = 1:2),
  by = NULL
)

if (exists("vigiar_rj_completude_tabela", mode = "function")) {
  vigiar_rj_completude_tabela(
    pm25_months,
    tabela = "df_mensal",
    anos_esperados = 2022L,
    meses_esperados = 1:3
  )
}
```

When `require_complete = TRUE`, incomplete coverage or possible truncation is an
error, not just a warning.

```{r, eval = FALSE}
vigiar_rj_completude_tabela(
  pm25_years,
  tabela = "df_anual",
  anos_esperados = 2020:2022,
  require_complete = TRUE
)
```

## Municipality-Specific Download

Use `vigiar_baixar_municipio()` when the target is one municipality. It filters
by normalized IBGE code, not by municipality name.

```{r, eval = FALSE}
campos <- vigiar_baixar_municipio("df_anual", codigo_ibge = 330100)

attr(campos, "vigiar_codigo_ibge_6")
attr(campos, "vigiar_codigo_ibge_7")
attr(campos, "vigiar_municipio")
```

## Online Download Example

The following code is intentionally not evaluated during vignette builds.

```{r, eval = FALSE}
vigiar_conectar()

pm25_rj <- vigiar_baixar_rj("df_anual", validar_cobertura = TRUE)
vigiar_rj_cobertura(pm25_rj)

pm25_rj_complete <- vigiar_baixar_rj(
  "df_anual",
  validar_cobertura = TRUE,
  require_complete = TRUE
)

vigiar_desconectar()
```

## Reproducible Online Audit

For a reproducible scientific audit, use `vigiar_auditar_rj_online()`. The
function writes `manifest.csv`, `manifest.json`, `audit.rds`, and per-table
coverage CSV files. These outputs include the table name, execution time,
package version, commit SHA when available, checksum, schema hash, row and
column counts, general coverage, table-grain completeness, missing
municipalities, incomplete health regions, truncation evidence, and an objective
conclusion.

```{r, eval = FALSE}
vigiar_conectar()

audit <- vigiar_auditar_rj_online(
  tabelas = c("df_anual", "df_mensal", "df_dias", "pop"),
  dominios_esperados = list(
    df_anual = list(anos_esperados = 2018:2023),
    df_mensal = list(anos_esperados = 2018:2023, meses_esperados = 1:12),
    df_dias = list(anos_esperados = 2018:2023, meses_esperados = 1:12),
    pop = list(anos_esperados = 2018:2023)
  ),
  salvar = TRUE,
  dir = "data-raw/rj-download-completeness-output",
  require_complete = TRUE
)

vigiar_desconectar()
```

The `require_complete = TRUE` mode turns partial coverage, possible truncation,
missing municipality codes, missing schema hashes, and failed downloads into a
clear error. It requires an explicit expected temporal domain. Use it for
release validation, not for exploratory checks.

## Possible Truncation

Large Power BI responses can be truncated. The package records
`vigiar_truncation_status` as `no_evidence`, `possible`, `probable`,
`confirmed`, or `unknown`, together with evidence. The raw response is assessed
before local filtering. Do not claim 92/92 completeness from a response with
truncation evidence without a validated partitioned download.

```{r}
attr(pm25_partial, "vigiar_possivel_truncamento") <- TRUE
coverage_fun(pm25_partial)
```

## Scientific Interpretation

Coverage tells you whether the expected municipalities appear in the downloaded
table. It does not prove that PM2.5 is measured at the ground, that the data are
fit for causal inference, or that the time series is sufficient for modelling.
The package does not implement GAM, DLNM, relative-risk, machine-learning, or
causal inference models. Treat coverage as a necessary diagnostic, not as a
scientific conclusion.
